\thetitle

### 2021-03-03 writing/discussion.tex
space war
pass over section 8

The analysis in the previous section suggests a number of interesting
high-level conclusions about blame in the gradual typing setting.  First,
run time type checks have a large positive impact. This is the case
regardless of whether these checks issue blame or throw plain exceptions.
Second, in many cases type checks that issue blame are more helpful than
those that do not. However, our results indicate that blame
is not critical in a majority of cases, and therefore it is worth further investigation whether
disabling blame tracking for the sake of performance may be   an
acceptable trade-off. Third, from the three different blame strategies, the
Natural approach of Typed Racket fares better than the Transient approach
of Reticulated Python, but only by a small margin. Since Natural offers complete
and sound path-based blame while Transient offers incomplete but sound heap-based
blame~\cite{gfd-oopsla-2019}, our results call for a deeper understanding
of the relative strengths of the two models for blame.  Fourth, given that
Transient's sound but shallow run time type checks do not seem to hamper
debugging, a version of Typed Racket that disables some of its wrappers
may offer an answer to the performance issues of Natural. 
Fifth, just as for Typed Racket, the experiment points to possible
improvements for Reticulated Python. In particular, the experimental 
data indicates that the usefulness of transient blame does not depend on
whether  the rational programmer follows first or the last element from a
blame set, and that using the last element seems to produce shorter trails.
Hence, given the performance issues of Transient we discuss further on,
a possible optimization is to limit the size of blame sets
to only contain the latest entries using timestamps.

There are a number of threats to the validity of our conclusions: (i) the
representativeness of benchmarks; (ii) the relation between mutations and
real programming mistakes; (iii) the definition of interesting debugging
scenarios; and (iv) our approach to sampling scenarios. The previous sections
have described the various ways we attempt to mitigate these threats;
we discuss two additional threats in the remainder of this section.

@@

Like {\em homo economicus\/}, which decouples the actual behavior of a
participant in an economy for the sake of mathematical modeling, the model of a
rational programmer decouples the actual debugging behavior of a software
developer for the sake of a systematic, large-scale analysis. This decoupling
comes with advantages and disadvantages. In the economic realm, mathematical models
have provided some predictive insights into the market's behavior; but as
behavioral economics has shown more recently, the mathematical abstraction of a
rational actor makes predictions also quite unreliable in some situations.
\footnote{It has also misled economists to focus on just the mathematics, though
this problem is not relevant here.}  Just like an ordinary consumer or producer,
an actual software developer is unlikely to stick to the exact strategy proposed
here. When this happens, the predicted benefits of blame assignment may not
materialize. Indeed, our own personal experience suggests such deviations, and
it also suggests that deviating is a mistake. To make a true judgment of the
usefulness of the rational-programmer idea, the community will need a lot more
experience with this form of evaluation and relating the evaluation to the
behavior of working programmers.

Relatedly, our setup hides the type choice a rational programmer must make. When the
run-time checks signal an impedance mismatch in the real world, a programmer
does not have a typed component ready to swap in. Instead, the programmer must
come up with the next set of types---and this means the programmer must make
choices. It is normally possible to assign consistent types to variables in a
component in different ways. The creation and curation of the benchmarks over
many years has driven home this lesson, but fortunately, it has also shown that
the types are in most cases reasonably canonical.  We therefore conjecture that
a rational programmer would in most cases come up with an equivalent type
assignment for trail extension as our experimental setup describes.

%% -----------------------------------------------------------------------------
\subsection{Threat: Can the Experiment Improve Transient?} \label{sec:threat:transient}

@@

An investigation of these empty blame cases reveals that Vitousek et al.'s
Transient blame assignment can lose track of the proper boundaries.  Despite
these losses, the number of failures is a small fraction of the overall number
of debugging scenarios. It is unlikely that any improvements to Vitousek et
al.'s strategy would change the overall conclusion about Transient blame.

{\bf Note} These scenarios suggest ways to improve Transient's blame map as
follows:
\begin{itemize}

@@

%% I don't see what this really means or contributes to the first two bullets:

 % \item The initial blame map must reflect the initial type environment.  For
 % instance, Typed Racket trusts that untyped core-library functions behave
 % correctly.

\end{itemize}
This small example is indicative of how rational-programmer experiments may
provide feedback.

@@

\item On large programs, Reticulated suffers from high overhead.  For example, a
Reticulated version of the simplest benchmark runs in ~40 seconds without blame,
and with blame times out after 10 minutes. Typed Racket runs the same benchmark
in a small fraction of the time.
\end{enumerate}
Future work on Transient blame is needed to make an informed decision. The
evaluation process presented here can help guide this research.

>>>>

The analysis in the previous section suggests a number of interesting
high-level conclusions about blame in the gradual typing setting.  First,
run time type checks have a large positive impact. This is the case
regardless of whether these checks issue blame or throw plain
exceptions.  Second, in many cases type checks that issue blame are more
helpful than those that do not. However, the results indicate that blame
is not critical in a majority of cases, and therefore it is worth further
investigation whether disabling blame tracking for the sake of performance
may be   an acceptable trade-off. Third, from the three different blame
strategies, the Natural approach of Typed Racket fares better than the
Transient approach of Reticulated Python, but only by a small margin.
Since Natural offers complete and sound path-based blame while Transient
offers incomplete but sound heap-based blame~\cite{gfd-oopsla-2019}, the
results call for a deeper understanding of the relative strengths of the
two models for blame.  Fourth, given that Transient's sound but shallow
run time type checks do not seem to hamper debugging, a version of Typed
Racket that disables some of its wrappers may offer an answer to the
well-known performance issues of Natural~\cite{gtnffvf-jfp-2019}.  Fifth,
just as for Typed Racket, the experiment points to possible improvements
for Reticulated Python. In particular, the experimental data indicates
that the usefulness of transient blame does not depend on whether  the
rational programmer follows first or the last element from a blame set,
and that using the last element seems to produce shorter trails.  Hence,
given the performance issues of Transient discussed further on, a possible
optimization is to limit the size of blame sets to only contain the latest
entries using timestamps.

There are a number of threats to the validity of these conclusions: (i) the
representativeness of benchmarks; (ii) the relation between mutations and
real programming mistakes; (iii) the definition of interesting debugging
scenarios; and (iv) the sampling strategy. The previous sections
have described the various ways we attempt to mitigate these threats;
the remainder of this section discussed three additional threats.

@@

Like {\em homo economicus\/}, which decouples the actual behavior of a
participant in an economy for the sake of mathematical modeling, the model of a
rational programmer decouples the actual debugging behavior of a software
developer for the sake of a systematic, large-scale analysis. This decoupling
comes with advantages and disadvantages. In the economic realm, mathematical models
have provided some predictive insights into the market's behavior; but as
behavioral economics has shown more recently, the mathematical abstraction of a
rational actor makes predictions also quite unreliable in some situations.
\footnote{It has also misled economists to focus on just the mathematics, though
this problem is not relevant here.}  Just like an ordinary consumer or producer,
an actual software developer is unlikely to stick to the exact strategy proposed
here. When this happens, the predicted benefits of blame assignment may not
materialize. Indeed, the authors's personal experience suggests such deviations, and
it also suggests that deviating is a mistake. To make a true judgment of the
usefulness of the rational-programmer idea, the community will need a lot more
experience with this form of evaluation and relating the evaluation to the
behavior of working programmers.

Relatedly, the experimental setup hides the type choice a rational programmer must make. When the
run-time checks signal an impedance mismatch in the real world, a programmer
does not have a typed component ready to swap in. Instead, the programmer must
come up with the next set of types---and this means the programmer must make
choices. It is normally possible to assign consistent types to variables in a
component in different ways. The creation and curation of the benchmarks over
many years has driven home this lesson, but fortunately, it has also shown that
the types are in most cases reasonably canonical.  The authors therefore conjecture that
a rational programmer would in most cases come up with an equivalent type
assignment in order to extend a trail as the experimental method describes.

%% -----------------------------------------------------------------------------
\subsection{Threat: Why Does Transient Lose Blame?} \label{sec:threat:transient}

@@

An investigation of these empty blame cases reveals that Transient is not
unsound but that Vitousek et al.'s Transient blame assignment loses track
of the proper boundaries due to an incomplete population of the blame map.
Despite these losses, the number of failures is a small fraction of the
overall number of debugging scenarios. It is unlikely that any
improvements to Vitousek et al.'s strategy would change the overall
conclusion about Transient blame.

{\bf Note} The scenarios where Transient returns empty blame 
suggest ways to improve Transient's blame map as follows:
\begin{itemize}

@@

\end{itemize}
This small example is indicative of how rational-programmer experiments may
provide actionable feedback for language design .

@@

\item On large programs, Reticulated suffers from high overhead.  For example, a
Reticulated version of the simplest benchmark runs in ~40 seconds without blame,
and with blame times out after 10 minutes. Typed Racket runs the same benchmark
in a small fraction of the time.
\end{enumerate}
More work on Transient blame is needed to make an informed decision.
